---
title: "Modular Programming"
author: "RStudyGroup; Shaurya"
date: "2018-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Today we will try and understand how modular programming works. Typically we want to make functions that do *ONE* thing really well. For today, let's try and make a simple temperature converter. It should let us convert **Fahrenheit <--> Celsius or Kelvin**.

---

Our first function only performs conversion of temperature from Fahrenheit to Celsius. That's it. Remember, `ONE thing, really well.`
```{r echo=TRUE, message=TRUE, warning=TRUE}

fahrenheit_to_celsius <- function(temperature_f) {
  # takes a temperature in Fahrenheit and converts it to Celsius
  temperature_c <- (temperature_f - 32) / 1.8
  return(temperature_c)
}
```
Our second function is to convert to Kelvin from Fahrenheit. Since this requires converting to Celsius first, we can reuse our code from above again.
```{r echo=TRUE, message=TRUE, warning=TRUE}

fahrenheit_to_kelvin <- function(temperature_f) {
  temperature_k <- fahrenheit_to_celsius(temperature_f) + 273.15
  return(temperature_k)
}
```
Finally, we make a GUI for calling both of our functions from above.
```{r }
gui <- function() {
  while(TRUE) { # allows us to keep going until user quits
    answer = readline(prompt = "Enter a temperature in Fahrenheit. Enter 'q' to quit. > ")
    
    # if the user wants to quit, check for 'q' first.
    if (identical(tolower(answer), 'q')) {
      break # breaks out of the while loop completely and exits
    }
    
    # convert the answer to a numeric since readline always takes everything as a character.
    answer <- as.numeric(answer)
    
    # if the answer is a long string, it will get converted to NA
    if(is.na(answer)) {
      return("Enter a number, dummy!")
      next # allows the while loop to start again.
    } else {
      # if the answer is a numeric value, ask for conversioin to either celsius or kelvin.
      answer2 <- tolower(readline(prompt = "Enter 'C' for celsius or 'K' for kelvin > "))
      if (identical(answer2, 'c')) {
        return(fahrenheit_to_celsius(answer))
      }
      if (identical(answer2, 'k')) {
        return(fahrenheit_to_kelvin(answer))
      } else {
        next
      }
    }
  }
}
```
