---
title: 'ComBEE RSG: Data Structures'
author: "M. Kartje"
date: "February 21, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Review of types

R has 5 atomic data types:
\vspace{12pt}

###Characters
```{r}
obj<-'Thursday'
obj
obj<-'This is also a character.'
obj
```
\vspace{12pt}

###Integers and Doubles (doubles sometimes called 'numeric')
```{r}
obj<-6
#example of a numeric
obj
typeof(obj)
obj<-as.integer(6)
#example of an integer
obj
typeof(obj)
```
\vspace{12pt}

###Logicals
```{r}
obj<-TRUE
#example of a logical (note lack of quotations in the definition)
obj
typeof(obj)
#OR
obj<-FALSE
obj
```
\vspace{12pt}

and complex numbers, which we won't discuss.

#Data Structures in R

R is an object-oriented language:

  + Data are stored in objects
  + Tasks are carried out by manipulating these objects


There are lots of different ways to arrange data in R.

Each specific way is called a 'data structure.'
\vspace{12pt}

##Vectors
The simplest type of data structure is called a `vector`.
```{r}
v_1<-c(1, 2, 3, 4, 5)
v_1
```

\vspace{12pt}

Try making a vector of characters. Name the vector `v_2`.
\vspace{12pt}

Run `typeof()` on the vectors `v_1` and `v_2`.
\vspace{12pt}

Try making a vector of some combination of numbers and characters.
\vspace{12pt}

In R, lots of functions are vectorized, meaning we can just pass the function or operation vectors, instead of passing individual scalars.

```{r}
6 + 6
```

Make a third vector containing the numbers 6 through 10. Name it `v_3`.
\vspace{12pt}

See what happens when you add this to `v_1` using the `+` operator. 
\vspace{12pt}

Repeat this with other operators:

  + Subtraction (`-`)
  + Multiplication (`*`)
  + blah blah you get the point.

\vspace{12pt}

###Indexing Vectors
The position of an entry in a vector is its _index_.

Revisiting the vector `v_3`:
```{r}
v_3
```

This vector has 5 elements. We can access individual elements using `v_3[]`.

```{r}
#Print the 3rd entry.
v_3[3]
```

Try printing one of the entries in your character vector `v_2`.
\vspace{12pt}


  
##Lists
Similarly to a `vector`, a `list` can be viewed as a sequence.

Unlike a `vector`, a `list` can support heterogeneous data types and can have >1 dimension.

```{r}
l_1<-list(1, 2, 3, 4)
l_1<-list(v_1)
```

  














