---
title: 'ComBEE RSG: Data Structures'
author: "M. Kartje"
date: "February 21, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Review of types

R has 5 atomic data types:
\vspace{12pt}

###Characters
```{r}
obj<-'Thursday'
obj
obj<-'This is also a character.'
obj
```
\vspace{12pt}

###Integers and Doubles (doubles sometimes called 'numeric')
```{r}
obj<-6
#example of a numeric
obj
typeof(obj)
obj<-as.integer(6)
#example of an integer
obj
typeof(obj)
```
\vspace{12pt}

###Logicals
```{r}
obj<-TRUE
#example of a logical (note lack of quotations in the definition)
obj
typeof(obj)
#OR
obj<-FALSE
obj
```
\vspace{12pt}

and complex numbers, which we won't discuss.

#Data Structures in R

R is an object-oriented language:

  + Data are stored in objects
  + Tasks are carried out by manipulating these objects


There are lots of different ways to arrange data in R.

Each specific way is called a 'data structure.'
\vspace{12pt}

##Vectors
The simplest type of data structure is called a `vector`.
```{r}
v_1<-c(1, 2, 3, 4, 5)
v_1
```

\vspace{12pt}

Try making a vector of characters. Name the vector `v_2`.
\vspace{12pt}

Run `typeof()` on the vectors `v_1` and `v_2`.
\vspace{12pt}

Try making a vector of some combination of numbers and characters.
\vspace{12pt}

In R, lots of functions are vectorized, meaning we can just pass the function or operation vectors, instead of passing individual scalars.

```{r}
6 + 6
```

Make a third vector containing the numbers 6 through 10. Name it `v_3`.
\vspace{12pt}

See what happens when you add this to `v_1` using the `+` operator. 
\vspace{12pt}

Repeat this with other operators:

  + Subtraction (`-`)
  + Multiplication (`*`)
  + blah blah you get the point.

\vspace{12pt}

###Indexing Vectors
The position of an entry in a vector is its _index_.

Revisiting the vector `v_3`:
```{r}
v_3<-c(6, 7, 8, 9, 10)
```

This vector has 5 elements. We can access individual elements using `v_3[]`.

```{r}
#print the 3rd entry.
v_3[3]
```

Try printing one of the entries in your character vector `v_2`.
\vspace{12pt}

Using `[]`, can also subset vectors.
```{r}
#print the 2nd - 5th entries
v_3[2:5]
```
\vspace{12pt}


##Lists
Similarly to a `vector`, a `list` can be viewed as a sequence.

Unlike a `vector`, a `list`:

  (1) can support heterogeneous data types
  (2) can have >1 dimension.

```{r}
l_1<-list(100, 54, 392, 47)
l_1

#print the first item of the list
l_1[1]
```

Make a list, `l_2`, from some combination types (_e.g._, numbers and characters).

W.R.T. bullet (1) above, how are different types combined into a list vs. a vector?

##Matricies
Matricies are simple, 2-dimensional data strucutres. 
R has lots of useful matrix functionality.
We probably won't use them too much.
\vspace{12pt}

```{r}
#a simple square matrix
m_1<-matrix(c(10, 9, 8,
              7, 6, 5,
              4, 3, 1), byrow = TRUE, nrow = 3)
m_1
```

The logic for subsetting matricies is similar two before, but with 2 dimensions `[ROW,COL]` .

To subset the entry in the 2nd row and 1st column, type

```{r}
m_1[2,1]
```

How might you subset an entire row or column?
\vspace{12pt}

##Dataframes
Dataframes are a super nifty data structure for data analysis.
  + 2-dimensional
  + Columns as variables
  + Rows as observations
  
```{r}
#a simple dataframe
df_1<-data.frame('col_1' = c(1, 2, 3, 4, 5), 'col_2' = c(FALSE, TRUE, TRUE, FALSE, TRUE), 'col_3' = c(10.7, 23.8, 98.7, 23.5, 99.7))
df_1
```

They can be indexed in a manner similar to matricies.

```{r}
df_1[2,3]
```

Alternatively, columns can be pulled out using `$`.

```{r}
df_1$row_2
```

Let's have a look at the `iris` dataframe.

```{r}
#look at first few rows
head(iris)

#examine the structure
str(iris)
```

By subsetting with `$` and using the `typeof()` function, determine which two other data strucutres dataframes are constructed from. 

The `mean()` and `sd()` functions compute the mean and standard deviation of a vector of numbers.

Using these functions, describe the iris dataset.
\vspace{24pt}

##Things to look forward to.

When properly organized into a dataframe, dataset can be very easy to summarize.
```{r}
boxplot(iris$Sepal.Length ~ iris$Species)
```
\vspace{24pt}

Data frames are also easy to manipulate in ways that are productive for data analysis.
```{r}
#subset to look only at setosa and versicolor
iris_sub<-subset(iris, Species == 'setosa' | Species == 'versicolor')
boxplot(iris_sub$Sepal.Length ~ iris_sub$Species)
t.test(iris$Sepal.Length[iris$Species == 'setosa'], iris$Sepal.Length[iris$Species == 'versicolor'])
```










