---
title: "RSG III: Data Visualization"
author: "M. Kartje"
date: "March 7, 2019"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Review of Data structures

I. Vectors

  + Homogeneous
  + 1-dimensional

EX  
```{r}
#a vector of numbers
vec_1<-c(1, 19, 34, 76)
vec_1
#a vector of characters
vec_2<-c('cat', 'dog', 'rat', 'pet')
vec_2
```


II. Lists

  + Heterogeneous
  + 1 - n-dimensional
  
EX
```{r}
#a list o numbers
l_1<-list(1, 19, 34, 76)
l_1
#a list of characters
l_2<-list('cat', 'dog', 'rat', 'pet')
l_2
#a list of lists
l_3<-list(l_1, l_2)
l_3
#lsit of vectors
```


III. Dataframes

  + List of vectors
  + 2-dimensional (matrix-like)
  + Homogeneous or heterogeneous?
  
EX
```{r}
#The iris dataframe
head(iris)
str(iris)
head(iris[,3])
head(iris$Petal.Length)

#The cars dataframe
head(cars)
str(cars)
head(cars[,2])
head(cars$dist)
```

\vspace{12pt}

Dataframes make plotting easy!

\vspace{12pt}

###Data Visualization

\vspace{12pt}

R has several basic functions for plotting data:

  + `hist()`
  + `plot()`
  + `boxplot()`

\vspace{12pt}

These functions are built in -- they come with R when you download it.

Various other plotting packages exist. 

By far, the most popular is ggplot2. 
We'll get to this today if there's time.


##Histograms

1) Look up how to use the `hist()` function using
`?hist()`

2) Use `hist()` to examine the frequency distribution of petal lengths in thie `iris` dataset.

```{r}
hist(iris$Petal.Length)
```

3) Subset the iris dataset to plot each species separately

\vspace{12pt}

4) Partition the plot window to view each of these plots

\vspace{12pt}

It looks like different species have different distributions of petal length...
How to best visualize this? 

\vspace{16pt}

##Boxplots

Boxplots allow for comparison of data across different levels of a factor.
\vspace{12pt}
Look at the help file for boxplots.
Using this info, construct a boxplot showing petal length vs. species

```{r}
#boxplot of petal length vs. species
boxplot(iris$Petal.Length ~ iris$Species)
```

1st, 2nd and 3rd quartiles (25, 50, 75 quantiles)

\vspace{12pt}

Let's look at another variable -- petal width

```{r}
#boxplot of petal width
boxplot(iris$Petal.Width ~ iris$Species)
```

\vspace{12pt}

The pattern across species seems similar, are these variables correlated in some way? 


##Scatterplots

The simplest scatterplots can be constructed using `plot()`.

Examine the `plot()` help file to see how it's used. What do you notice about the way arguments can be supplied?

```{r}
#scatter plot of petal width vs petal length
plot(iris$Petal.Length, iris$Petal.Width)
#add a regression line for fun
abline(lm(iris$Petal.Width ~ iris$Petal.Length))
```

#Brief stats foray.

#install ggplot2.
